Length Gauge Theory of Surface Second Harmonic Generation

### 2014-11-20 paper-v-0.tex
fixing bib file and cleaning up paper

\documentclass[floatfix,prb,aps,superscriptaddress,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\allowdisplaybreaks[1]%biutiful equation breaker!!!
\usepackage{ulem}
\usepackage{subfigure}
%\input{/Users/bms/util/definitions}
%\def\chon[1]{\color{red} #1}
\def\chon{black}
\def\chonn{red}
\def\nicolas{red}
\input{definitions}
\usepackage{bm}
%%%%*****************%%%% fine hyperef 
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{showlabels}
%\usepackage{showkeys}

@@

\begin{abstract}
We present a theoretical review of surface second harmonic generation (SHG) 
from semiconductor surfaces based on the {\color{\chon} length} gauge. This 
layer-by-layer analysis is carefully presented in order to show how a 
surface SHG calculation can be readily evaluated. The nonlinear susceptibility 
tensor for a surface, $\boldsymbol{\chi}^S(-2\go;\go,\go)$
is split into two terms relating to inter-band 
and intra-band one-electron transitions.

@@

In recent years, surface nonlinear optical spectroscopies and in 
particular, surface second-harmonic generation SSHG have evolved as useful 
nondestructive and noninvasive tool to study properties of surfaces 
and interfaces: atomic structure, phase transitions, and adsorption of 
atoms, among others.\cite{
    daumPRL93, 
    mcgilpOE94, 
    meyerPRL95, 
    powerPRL95, 
    godefroyAPL96, 
    hoferAPA96, 
    dadapPRB97, 
    bloembergenAPB99, 
    mcgilpSRL99, 
    suzukiAPB99, 
    mitchellSS01,
    hughesPRB96, 
    guyot-sionnestPRB88, 
    downerPSSA01, 
    shenAPB99, 
    shenNAT89, 
    chenPRL81, 
    mendozaPRL98,
    downerSIA01,
    cazzanelliNM14}
Nowadays, SSHG spectroscopy is a crucial tool for research and development
in microelectronics,\cite{zheltikovLP00} semiconductors,
\cite{lupkeSSR99}, nanomaterials,\cite{salazar-aparicioPRB14} 
and many more recent 
areas of great scientific and commercial interest.\cite{cazzanelliNM14}
The high surface sensitivity of SSHG spectroscopy is due to the
fact that, within the dipole approximation, the bulk SHG signal of
centrosymmetric materials is identically zero, and thus only the
surface, where the inversion symmetry is broken, can radiate. 
Much of the foundation of surface science has been built from
experiments involving emission or scattering of electrons from
surfaces. These require ultrahigh vacuum (UHV) environments and
provide no access to buried interfaces. 
However, SSHG
is compatibile with non-UHV ambients and has access to
interfaces buried beneath transparent overlayers. Even when applied to
surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

@@

As it is well known
Density Functional Theory (DFT) within the Local Density Approximation
(LDA),\cite{kohnPR65} the most 
used framework for \textit{ab initio} calculations,
underestimates the energy band gap of semiconductors. It is also well
understood that one has to include the many-body interaction to
correct for the underestimation of the gap. In this context the so
called GW approximation\cite{onidaRMP02} is known to correct the band
gap of most semiconductors\cite{luceroJPCM12}. However this could be a numerically very
expensive calculation, and thus one uses the much simpler scissors
operator scheme.%%\cite{?} find this
This allows to ``open'' the DFT-LDA gap to
its correct experimental or GW value, when experiments are not available. 
In this context,
to correct for the underestimation of the energy band gap of semiconductors, 
Nastos et al.\cite{nastosPRB05} 
using the 
``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' 
showed how to correctly include in the SH susceptibility the many-body corrections through the
scissors operator.
Later,
Cabellos \textit{et al}.\cite{cabellosPRB09} 
elaborated a derivation of the ``velocity gauge'' or
``$\bfA\cdot\bfv$'' gauge properly including the scissors
operator, and proved  gauge 
invariance with respect to the length gauge.
From these works, it is clear the length gauge is a much better staring
point to obtain the surface SH (SSH) susceptibility, as it will be
elaborated in this article. Above only being valid for bulk semiconductors.

@@

We use the IPA and neglect local field  
and excitonic effects{\color{\chon}.  
We assume a classical} electromagnetic field
{\color{\chon} and quantum mechanical matter.}
We can describe the system using 
%a scaled 
the one electron density
operator ${\rho}$, with which we can calculate the expectation value of a
single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle=\mbox{Tr}({\rho} {\cal O}),  
\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr
the trace. 
The density operator satisfies
$
i\hbar (d{\rho}/dt)=[H(t),{\rho}],  
\label{rho}
$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t)=H_{0}+H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the
electron with the electromagnetic field.
% ; $H_{0}$ has eigenvalues
% $\hbar \omega_{n}(\mathbf{k})$
% and eigenstates $| n\mathbf{k} \rangle$ (Bloch states) labeled by a band 
% index $n$ and crystal momentum $\mathbf{k}$.
To proceed with the solution of $\rho$  it is convenient to use the
interaction picture, where a unitary operator $U=\exp ({iH_{0}t/\hbar })$
transforms any operator $\mathcal{O}$ into 
$\tilde{\mathcal{O}}=U{\cal O}U^{\dagger }$. Even if $\mathcal{O}$ 
{\color{\chon} is time-independent}, 
$\tilde{\mathcal{O}}$ {\color{\chon} is time-dependent} through the explicit
time dependence of $U$. 
The dynamical 
equation for $\tilde{\rho}$ is 
given by
\begin{equation*}
i\hbar \frac{d\tilde{{\rho}}}{dt}=[\tilde{H}_{I}(t),\tilde{{\rho}}],  
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{{\rho}}(t)=i\hbar \tilde{{\rho}}_{0}+\int_{-\infty }^{t}dt^{\prime }[
\tilde{H}_{I}(t^{\prime }),\tilde{\rho}(t^{\prime })],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0}=\tilde{\rho}(t=-\infty )$ is the unperturbed density matrix. We look
for the standard perturbation series solution, 
$\tilde{\rho}(t)=\tilde{\rho}^{(0)}+\tilde{\rho}^{(1)}+\tilde{\rho}^{(2)}+\cdots$,  
where the superscript denotes the order (power) with which each term depends
on the perturbation $H_{I}(t)$. From Eq.~\eqref{trans} the $N$-th order
term is 
\begin{equation}
\tilde{{\rho}}^{(N)}(t)=\frac{1}{i\hbar }\int_{-\infty }^{t}dt^{\prime }[\tilde{
H}_{I}(t^{\prime }),\tilde{\rho}^{(N-1)}(t^{\prime })].  
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}
^{(0)}\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle=f(\hbar \omega_{n}(\mathbf{k}))\equiv f_{n}$. For a
clean, cold semiconductor $f_{n}=1$ {\color{\chon} when} $n$ {\color{\chon} is} a valence ($v$) or
occupied band, and zero {\color{\chon} when} $n$ {\color{\chon} is} a conduction ($c$) or empty band.
{\color{\chon} We assume this for the remainder of the article.}
{\color{\chon} T}he expectation values must satisfy
$
\langle{\cal O}\rangle=
\mbox{Tr}({\rho}{\cal O})
=
\mbox{Tr}(\tilde{{\rho}}\tilde{\cal O})
$.

We will look for the expectation value of the microscopic current density, 
$\mathbf{J}$, given by 
\begin{equation}
\bfJ=\langle{\mathbf{J}}\rangle=\frac{e}{A}\mbox{Tr}({\rho}\dot{\mathbf{r}})
,
\label{pe}
\end{equation}
where $\dot{\mathbf{r}}$ is the time derivative of the position operator of the
electron {\color{\chon} with} charge $e$, {\color{\chon} defined as}
\begin{equation}
\mathbf{v}\equiv \dot{\mathbf{r}}=\frac{1}{i\hbar }[\mathbf{r},H_0],  
\label{mv}
\end{equation}
with $\mathbf{v}$ the velocity operator of the electron, and $A$ the
area of the unit cell. We calculate the macroscopic polarization density 
$\mathbf{P}$, related to $\bfJ$ by
$\bfJ=d\mathbf{P}/dt$. For a 
perturbing (Maxwell macroscopic) electromagnetic field, $\mathbf{E}(t)=
\mathbf{E}(\omega )e^{-i\tilde{\omega} t}+c.c.$,
where $\tilde\omega=\omega+i\eta $,
and $\eta >0$ {\color{\chon} adiabatically switches on} the interaction,
we write the second order nonlinear
polarization as, 
\begin{equation}
\calp^a(2\omega)=\chi ^{abc}(-2\omega;\omega,\omega)E^{b}(\omega)E^{c}(\omega),  
\label{pshg}
\end{equation}
where $\chi^{abc}(-2\omega ;\omega ,\omega )$ is the nonlinear
susceptibility responsible {\color{\chon} for} surface second harmonic generation
(SSHG). 
The 
superscripts in Eq.~\eqref{pshg} denote Cartesian components, and if
repeated are to be summed over. Without loss of generality we can always
define $\chi^{abc}(-2\omega;\omega,\omega)$
 to satisfy intrinsic permutation
symmetry, 
$\chi^{abc}(-2\omega ;\omega ,\omega )=\chi ^{acb}(-2\omega ;\omega
,\omega )$.
% ;
% if it did not, the part that did not satisfy intrinsic permutation symmetric
% would make no contribution to the second order polarization. This part
% could be dropped since it would
% have no physical significance.

The unperturbed Hamiltonian 
is used to solve the Kohn-Sham equations\cite{kohnPR65} of Density  
Functional Theory (DFT) {\color{\chon}. It is convenient to work within the Local 
Density Approximation (LDA), so we label the hamiltonian with the corresponding  
LDA superscript. Any other approximation can be used (like GGA) and our 
derivation remains the same.} Then,
\begin{equation}\label{h0}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)
=\frac{p^{2}}{2m_e}+V(\bfr,\bfp)
\end{equation}
with $m_e$ the mass of the electron, $\mathbf{p}$ its canonical momentum, and 
$V$
 the periodic crystal potential, where we neglect spin-orbit terms.
To be more general in our derivation of
$\chi^{abc}(-2\omega;\omega,\omega)$, we asume
that the crystal potential could have a {\color{\chon} non-local spatial}
contribution {\color{\chon} as} is customary for most
pseudopotentials, and then we replace $V$ with
\begin{align}\label{ps}
V^{\mathrm{ps}}(\bfr,\bfp)=V(\bfr)+V^{\mathrm{nl}}(\bfr,\bfp)
,
\end{align}
where 
$V(\bfr)$ and $V^{\mathrm{nl}}(\bfr,\bfp)$  are the local and 
non-local contributions to the pseudopotential $V^{\mathrm{ps}}$,
 respectively.
In case of
a local potential, i.e. $V= V(\bfr)$, 
like that of all-electron schemes, 
we simply omit the contribution coming
from the non-local part of the pseudopotential, $V^{\mathrm{nl}}$, in
the results that we {\color{\chon} have derived.}

{\color{\chon} It} is well known that the use of the {\color{\chon} LDA} 
leads to an underestimation of the band gap. A standard
procedure to correct for this is to
use {\color{\chon} the} ``scissors approximation'', {\color{\chon} where the 
conduction bands are rigidly shifted in energy} so that the band gap 
corresponds to the accepted experimental band gap{\color{\chon} .
This} is often in fairly good agreement with the GW
band gap based on a more sophisticated calculation.\cite{hybertsenPRB86}
{\color{\chon} T}he LDA wave functions are used since they produce band
structures with dispersion relations similar to those predicted by the GW
approximation. Mathematically, {\color{\chon} the scissors (non-local) operator 
$S$ is added} to the unperturbed or unscissored Hamiltonian {\color{\chon} $H^{\mathrm{LDA}}_{0}$ ,}
\begin{equation*}
H^\sigma_{0}(\bfr,\bfp)=H^{\mathrm{LDA}}_{0}(\bfr,\bfp)+S(\bfr,\bfp)
\end{equation*}
where 
\begin{equation}
S(\bfr,\bfp)=\hbar \sigma\sum_{n}\int d^{3}k(1-f_{n})
|n\mathbf{k}\rangle\langle n\mathbf{k}|,
\label{hats}
\end{equation}
with $\hbar \sigma$  the rigid ($\mathbf{k}$-independent) energy correction to be
applied. 
%Several properties of $S(\mathbf{r},\mathbf{p})$ are shown in the
%appendix.
The unscissored and scissored Hamiltonians satisfy 
\begin{eqnarray*}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega^{\mathrm{LDA}}_{n}(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\label{hamils} \\
H_{0}^\sigma (\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega_{n}^\sigma
(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\end{eqnarray*}
where the scissor-shifted energies, 
$\omega_{n}^\sigma(\mathbf{k})$ , are given by
\begin{equation}
\omega_{n}^\sigma(\mathbf{k})=\omega^\lda_{n}(\mathbf{k})+(1-f_{n})\sigma.
\label{wese}
\end{equation}
We emphasize that the {\color{\chon} scissored and unscissored Hamiltonian 
have the same eigenfunctions,} where
$\psi_{n\bfk}(\bfr)=\braket{\bfr}{n\bfk}=e^{i\bfk\cdot\bfr}u_{n\bfk}(\bfr)$,
are the real space representations of the Bloch states $\ket{n\bfk}$ labelled 
by the band index $n$ and the crystal momentum $\bfk$, and $u_{n\bfk}(\bfr)$
is cell periodic.

\subsection{Length Gauge Formalism}
To calculate the optical response 
within the Length Gauge {\color{\chon} formalism}, we take the 
interaction Hamiltonian {\color{\chon} as} 
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).  \label{rde}
\end{equation}
The 
treatment of the position operator
$\bfr$ for extended Bloch states, $\braket{\bfr}{n\bfk}$ {\color{\chon} has} 
been described {\color{\chon} in detail}.\cite{adamsJCP53,blountSSP62} 
{\color{\chon} In particular, Ref.~\onlinecite{aversaPRB95} develops} 
the correct treatment of $\bfr$ for
the calculation of the {\color{\chon} nonlinear} optical {\color{\chon} response.} 
$\bfr$ is split into the {\it intraband} 
($\bfr_i$) and {\it interband} ($\bfr_e$) parts, where 
$\bfr=\bfr_i+\bfr_e$. 
For $\bfr_e$ one uses
\begin{align}\label{rnminn}
\bra{n\bfk}\bfr_e\ket{m\bfk'} &=
(1-\gd_{nm})\gd(\bfk-\bfk')\bfgxi_{nm}(\bfk) 
,
\end{align}
such that $\bfr_{e,nm}=0$ for $n=m${\color{\chon} .
From Eq.~\eqref{mv}} with $H_0\to
H^\sigma_0$, we obtain
\begin{align}\label{pmnrmn}
\bfr_{e,nm}(\bfk) =
\bfgxi_{nm}(\bfk)\equiv 
\bfr_{nm}(\bfk) 
&=
\frac{\bfv^\gs_{nm}(\bfk)}{i\go^\gs_{nm}(\bfk)}
\quad\quad n\notin D_m 
,
\end{align}  
where we defined
$\go^\gs_{nm}(\bfk)\equiv\go^\gs_n(\bfk)-\go^\gs_m(\bfk)$, and
$D_m$ are all the possible degenerate $m$-states. 
%We recognize that for each instance in which 
%$n\notin D_m$ or $n\ne m$, $\bfr_{nm}(\bfk)$ are interband matrix
%elements.
For the intraband part, {\color{\chon} $\bfr_i$ only appears in
commutators during the derivation of
the optical response. We use},\cite{aversaPRB95}
\begin{equation}\label{conmri3n}
\bra{n\bfk}[\bfr_i,\calo]\ket{m\bfk'}
=i\gd(\bfk-\bfk')(\calo_{nm})_{;\bfk}
,
\end{equation}  
where
\begin{equation}\label{gendevnn}
(\calo_{nm})_{;\bfk}=
\nabla_\bfk 
\calo_{n m}(\bfk) 
- 
i 
\calo_{nm}(\bfk) 
\left(
\bfgxi_{nn}(\bfk) 
-
\bfgxi_{mm}(\bfk) 
\right) 
,
\end{equation} 
is
the generalized derivative 
of the operator $\calo$. 
The vectors $\bfgxi_{nn}(\bfk)$ are defined in 
Ref.~\onlinecite{aversaPRB95} {\color{\chon} though} they do not need to be 
calculated explicitly in what follows.

>>>>

\documentclass[floatfix,prb,aps,superscriptaddress,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\allowdisplaybreaks[1]%biutiful equation breaker!!!
\usepackage{ulem}
\usepackage{subfigure}
%\input{/Users/bms/util/definitions}
%\def\chon[1]{\color{red} #1}
\def\chon{red}
\def\chonn{red}
\def\nicolas{red}
\input{definitions}
\usepackage{bm}
%%%%*****************%%%% fine hyperef 
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{showlabels}
%\usepackage{showkeys}

@@

\begin{abstract}
We present a theoretical review of surface second harmonic generation (SHG) 
from semiconductor surfaces based on the length gauge. This 
layer-by-layer analysis is carefully presented in order to show how a 
surface SHG calculation can be readily evaluated. The nonlinear susceptibility 
tensor for a surface, $\boldsymbol{\chi}^S(-2\go;\go,\go)$
is split into two terms relating to inter-band 
and intra-band one-electron transitions.

@@

In recent years, surface nonlinear optical spectroscopies and in 
particular, surface second-harmonic generation SSHG have evolved as useful 
nondestructive and noninvasive tool to study properties of surfaces 
and interfaces: atomic structure, phase transitions, and adsorption of 
atoms, among others.\cite{
    daumPRL93, 
    mcgilpOE94, 
    meyerPRL95, 
    powerPRL95, 
    godefroyAPL96, 
    hoferAPA96, 
    dadapPRB97, 
    bloembergenAPB99, 
    mcgilpSRL99, 
    suzukiAPB99, 
    mitchellSS01,
    hughesPRB96, 
    guyotPRB88, 
    downerPSSA01, 
    shenAPB99, 
    shenNAT89, 
    chenPRL81, 
    mendozaPRL98,
    downerSIA01,
    cazzanelliNM14}
Nowadays, SSHG spectroscopy is a crucial tool for research and development
in microelectronics,\cite{zheltikovLP00} semiconductors,
\cite{lupkeSSR99}, nanomaterials,\cite{salazar-aparicioPRB14} 
and many more recent 
areas of great scientific and commercial interest.\cite{cazzanelliNM14}
The high surface sensitivity of SSHG spectroscopy is due to the
fact that, within the dipole approximation, the bulk SHG signal of
centrosymmetric materials is identically zero, and thus only the
surface, where the inversion symmetry is broken, can radiate. 
Much of the foundation of surface science has been built from
experiments involving emission or scattering of electrons from
surfaces. These require ultrahigh vacuum (UHV) environments and
provide no access to buried interfaces. 
However, SSHG
is compatibile with non-UHV ambients and has access to
interfaces buried beneath transparent overlayers. Even when applied to
surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

@@

As it is well known
Density Functional Theory (DFT) within the Local Density Approximation
(LDA),\cite{kohnPR65} the most 
used framework for \textit{ab initio} calculations,
underestimates the energy band gap of semiconductors. It is also well
understood that one has to include the many-body interaction to
correct for the underestimation of the gap. In this context the so
called GW approximation\cite{onidaRMP02} is known to correct the band
gap of most semiconductors\cite{luceroJPCM12}. However this could be a numerically very
expensive calculation, and thus one uses the much simpler scissors
operator scheme.\cite{levinePRL89, levinePRL91}
This allows to ``open'' the DFT-LDA gap to
its correct experimental or GW value, when experiments are not available. 
In this context,
to correct for the underestimation of the energy band gap of semiconductors, 
Nastos et al.\cite{nastosPRB05} 
using the 
``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' 
showed how to correctly include in the SH susceptibility the many-body corrections through the
scissors operator.
Later,
Cabellos \textit{et al}.\cite{cabellosPRB09} 
elaborated a derivation of the ``velocity gauge'' or
``$\bfA\cdot\bfv$'' gauge properly including the scissors
operator, and proved  gauge 
invariance with respect to the length gauge.
From these works, it is clear the length gauge is a much better staring
point to obtain the surface SH (SSH) susceptibility, as it will be
elaborated in this article. Above only being valid for bulk semiconductors.

@@

We use the IPA and neglect local field  
and excitonic effects.  
We assume a classical electromagnetic field
and quantum mechanical matter.
We can describe the system using 
%a scaled 
the one electron density
operator ${\rho}$, with which we can calculate the expectation value of a
single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle=\mbox{Tr}({\rho} {\cal O}),  
\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr
the trace. 
The density operator satisfies
$
i\hbar (d{\rho}/dt)=[H(t),{\rho}],  
\label{rho}
$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t)=H_{0}+H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the
electron with the electromagnetic field.
% ; $H_{0}$ has eigenvalues
% $\hbar \omega_{n}(\mathbf{k})$
% and eigenstates $| n\mathbf{k} \rangle$ (Bloch states) labeled by a band 
% index $n$ and crystal momentum $\mathbf{k}$.
To proceed with the solution of $\rho$  it is convenient to use the
interaction picture, where a unitary operator $U=\exp ({iH_{0}t/\hbar })$
transforms any operator $\mathcal{O}$ into 
$\tilde{\mathcal{O}}=U{\cal O}U^{\dagger }$. Even if $\mathcal{O}$ 
is time-independent, 
$\tilde{\mathcal{O}}$ is time-dependent through the explicit
time dependence of $U$. 
The dynamical 
equation for $\tilde{\rho}$ is 
given by
\begin{equation*}
i\hbar \frac{d\tilde{{\rho}}}{dt}=[\tilde{H}_{I}(t),\tilde{{\rho}}],  
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{{\rho}}(t)=i\hbar \tilde{{\rho}}_{0}+\int_{-\infty }^{t}dt^{\prime }[
\tilde{H}_{I}(t^{\prime }),\tilde{\rho}(t^{\prime })],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0}=\tilde{\rho}(t=-\infty )$ is the unperturbed density matrix. We look
for the standard perturbation series solution, 
$\tilde{\rho}(t)=\tilde{\rho}^{(0)}+\tilde{\rho}^{(1)}+\tilde{\rho}^{(2)}+\cdots$,  
where the superscript denotes the order (power) with which each term depends
on the perturbation $H_{I}(t)$. From Eq.~\eqref{trans} the $N$-th order
term is 
\begin{equation}
\tilde{{\rho}}^{(N)}(t)=\frac{1}{i\hbar }\int_{-\infty }^{t}dt^{\prime }[\tilde{
H}_{I}(t^{\prime }),\tilde{\rho}^{(N-1)}(t^{\prime })].  
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}
^{(0)}\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle=f(\hbar \omega_{n}(\mathbf{k}))\equiv f_{n}$. For a
clean, cold semiconductor $f_{n}=1$ when $n$ is a valence ($v$) or
occupied band, and zero when $n$ is a conduction ($c$) or empty band.
We assume this for the remainder of the article.
The expectation values must satisfy
$
\langle{\cal O}\rangle=
\mbox{Tr}({\rho}{\cal O})
=
\mbox{Tr}(\tilde{{\rho}}\tilde{\cal O})
$.

We will look for the expectation value of the microscopic current density, 
$\mathbf{J}$, given by 
\begin{equation}
\bfJ=\langle{\mathbf{J}}\rangle=\frac{e}{A}\mbox{Tr}({\rho}\dot{\mathbf{r}})
,
\label{pe}
\end{equation}
where $\dot{\mathbf{r}}$ is the time derivative of the position operator of the
electron with charge $e$, defined as
\begin{equation}
\mathbf{v}\equiv \dot{\mathbf{r}}=\frac{1}{i\hbar }[\mathbf{r},H_0],  
\label{mv}
\end{equation}
with $\mathbf{v}$ the velocity operator of the electron, and $A$ the
area of the unit cell. We calculate the macroscopic polarization density 
$\mathbf{P}$, related to $\bfJ$ by
$\bfJ=d\mathbf{P}/dt$. For a 
perturbing (Maxwell macroscopic) electromagnetic field, $\mathbf{E}(t)=
\mathbf{E}(\omega )e^{-i\tilde{\omega} t}+c.c.$,
where $\tilde\omega=\omega+i\eta $,
and $\eta >0$ adiabatically switches on the interaction,
we write the second order nonlinear
polarization as, 
\begin{equation}
\calp^a(2\omega)=\chi ^{abc}(-2\omega;\omega,\omega)E^{b}(\omega)E^{c}(\omega),  
\label{pshg}
\end{equation}
where $\chi^{abc}(-2\omega ;\omega ,\omega )$ is the nonlinear
susceptibility responsible for surface second harmonic generation
(SSHG). 
The 
superscripts in Eq.~\eqref{pshg} denote Cartesian components, and if
repeated are to be summed over. Without loss of generality we can always
define $\chi^{abc}(-2\omega;\omega,\omega)$
 to satisfy intrinsic permutation
symmetry, 
$\chi^{abc}(-2\omega ;\omega ,\omega )=\chi ^{acb}(-2\omega ;\omega
,\omega )$.
% ;
% if it did not, the part that did not satisfy intrinsic permutation symmetric
% would make no contribution to the second order polarization. This part
% could be dropped since it would
% have no physical significance.

The unperturbed Hamiltonian 
is used to solve the Kohn-Sham equations\cite{kohnPR65} of Density  
Functional Theory (DFT). It is convenient to work within the Local 
Density Approximation (LDA), so we label the hamiltonian with the corresponding  
LDA superscript. Any other approximation can be used (like GGA) and our 
derivation remains the same. Then,
\begin{equation}\label{h0}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)
=\frac{p^{2}}{2m_e}+V(\bfr,\bfp)
\end{equation}
with $m_e$ the mass of the electron, $\mathbf{p}$ its canonical momentum, and 
$V$
 the periodic crystal potential, where we neglect spin-orbit terms.
To be more general in our derivation of
$\chi^{abc}(-2\omega;\omega,\omega)$, we asume
that the crystal potential could have a non-local spatial
contribution as is customary for most
pseudopotentials, and then we replace $V$ with
\begin{align}\label{ps}
V^{\mathrm{ps}}(\bfr,\bfp)=V(\bfr)+V^{\mathrm{nl}}(\bfr,\bfp)
,
\end{align}
where 
$V(\bfr)$ and $V^{\mathrm{nl}}(\bfr,\bfp)$  are the local and 
non-local contributions to the pseudopotential $V^{\mathrm{ps}}$,
 respectively.
In case of
a local potential, i.e. $V= V(\bfr)$, 
like that of all-electron schemes, 
we simply omit the contribution coming
from the non-local part of the pseudopotential, $V^{\mathrm{nl}}$, in
the results that we have derived.

It is well known that the use of the LDA
leads to an underestimation of the band gap. A standard
procedure to correct for this is to
use the ``scissors approximation'', where the 
conduction bands are rigidly shifted in energy so that the band gap 
corresponds to the accepted experimental band gap.
This is often in fairly good agreement with the GW
band gap based on a more sophisticated calculation.\cite{hybertsenPRB86}
The LDA wave functions are used since they produce band
structures with dispersion relations similar to those predicted by the GW
approximation. Mathematically, the scissors (non-local) operator 
$S$ is added to the unperturbed or unscissored Hamiltonian $H^{\mathrm{LDA}}_{0}$ ,
\begin{equation*}
H^\sigma_{0}(\bfr,\bfp)=H^{\mathrm{LDA}}_{0}(\bfr,\bfp)+S(\bfr,\bfp)
\end{equation*}
where 
\begin{equation}
S(\bfr,\bfp)=\hbar \sigma\sum_{n}\int d^{3}k(1-f_{n})
|n\mathbf{k}\rangle\langle n\mathbf{k}|,
\label{hats}
\end{equation}
with $\hbar \sigma$  the rigid ($\mathbf{k}$-independent) energy correction to be
applied. 
%Several properties of $S(\mathbf{r},\mathbf{p})$ are shown in the
%appendix.
The unscissored and scissored Hamiltonians satisfy 
\begin{eqnarray*}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega^{\mathrm{LDA}}_{n}(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\label{hamils} \\
H_{0}^\sigma (\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega_{n}^\sigma
(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\end{eqnarray*}
where the scissor-shifted energies, 
$\omega_{n}^\sigma(\mathbf{k})$ , are given by
\begin{equation}
\omega_{n}^\sigma(\mathbf{k})=\omega^\lda_{n}(\mathbf{k})+(1-f_{n})\sigma.
\label{wese}
\end{equation}
We emphasize that the scissored and unscissored Hamiltonian 
have the same eigenfunctions, where
$\psi_{n\bfk}(\bfr)=\braket{\bfr}{n\bfk}=e^{i\bfk\cdot\bfr}u_{n\bfk}(\bfr)$,
are the real space representations of the Bloch states $\ket{n\bfk}$ labelled 
by the band index $n$ and the crystal momentum $\bfk$, and $u_{n\bfk}(\bfr)$
is cell periodic.

\subsection{Length Gauge Formalism}
To calculate the optical response 
within the Length Gauge formalism, we take the 
interaction Hamiltonian as
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).  \label{rde}
\end{equation}
The 
treatment of the position operator
$\bfr$ for extended Bloch states, $\braket{\bfr}{n\bfk}$ has 
been described in detail.\cite{adamsJCP53,blountSSP62} 
In particular, Ref.~\onlinecite{aversaPRB95} develops
the correct treatment of $\bfr$ for
the calculation of the nonlinear optical response.
$\bfr$ is split into the {\it intraband} 
($\bfr_i$) and {\it interband} ($\bfr_e$) parts, where 
$\bfr=\bfr_i+\bfr_e$. 
For $\bfr_e$ one uses
\begin{align}\label{rnminn}
\bra{n\bfk}\bfr_e\ket{m\bfk'} &=
(1-\gd_{nm})\gd(\bfk-\bfk')\bfgxi_{nm}(\bfk) 
,
\end{align}
such that $\bfr_{e,nm}=0$ for $n=m$.
From Eq.~\eqref{mv} with $H_0\to
H^\sigma_0$, we obtain
\begin{align}\label{pmnrmn}
\bfr_{e,nm}(\bfk) =
\bfgxi_{nm}(\bfk)\equiv 
\bfr_{nm}(\bfk) 
&=
\frac{\bfv^\gs_{nm}(\bfk)}{i\go^\gs_{nm}(\bfk)}
\quad\quad n\notin D_m 
,
\end{align}  
where we defined
$\go^\gs_{nm}(\bfk)\equiv\go^\gs_n(\bfk)-\go^\gs_m(\bfk)$, and
$D_m$ are all the possible degenerate $m$-states. 
%We recognize that for each instance in which 
%$n\notin D_m$ or $n\ne m$, $\bfr_{nm}(\bfk)$ are interband matrix
%elements.
For the intraband part, $\bfr_i$ only appears in
commutators during the derivation of
the optical response. We use,\cite{aversaPRB95}
\begin{equation}\label{conmri3n}
\bra{n\bfk}[\bfr_i,\calo]\ket{m\bfk'}
=i\gd(\bfk-\bfk')(\calo_{nm})_{;\bfk}
,
\end{equation}  
where
\begin{equation}\label{gendevnn}
(\calo_{nm})_{;\bfk}=
\nabla_\bfk 
\calo_{n m}(\bfk) 
- 
i 
\calo_{nm}(\bfk) 
\left(
\bfgxi_{nn}(\bfk) 
-
\bfgxi_{mm}(\bfk) 
\right) 
,
\end{equation} 
is
the generalized derivative 
of the operator $\calo$. 
The vectors $\bfgxi_{nn}(\bfk)$ are defined in 
Ref.~\onlinecite{aversaPRB95} though they do not need to be 
calculated explicitly in what follows.
